# Reproducible Research: Peer Assessment 1

## Tools
Here I include constants and routines that are used in the rest of the document.

#### Constants
```{r}
DATA_FILE = "..\\..\\Data\\activity.csv"
```

#### Support functions
```{r}
# Converts a date (in text) and time of numeric format to POSIXct format
#   Input: dt_text is the date (text format, yyyy-mm-dd)
#   Input: tm_number is the time (integer of form hM, e.g. 105 is 1:05)
#   Output: Time/Date stamp, combining the two pieces of information
convert_date_time = function(dt_text,tm_number) {
    tm_text = sprintf("%04d",tm_number)
    date_time_text = paste(dt_text, tm_text)
    result = as.POSIXct(date_time_text,format = "%Y-%m-%d %H%M", tz="GMT")
    return(result) #I prefer explicit, simple returns - not efficient but...
}

#Test; should return 2012-12-14 01:20:00 GMT
convert_date_time("2012-12-14",120)
```

## Loading and preprocessing the data

#### Load
```{r load}
dta = read.csv(DATA_FILE)
```
#### Process (Add *date_time* and *time* columns to *dta*)
```{r add times}
dta$date_time = mapply(convert_date_time,
                       dta$date,
                       dta$interval,
                       SIMPLIFY=FALSE
                       )
dta$time = unlist(mapply(function(a){unlist(format(a,"%H:%M"))}, 
                         dta$date_time, 
                         SIMPLIFY=FALSE
                         )
                  )
dta$time = as.POSIXct(dta$time, format="%H:%M")
```

## What is mean total number of steps taken per day?
```{r steps_evaluation}
with(dta,hist(steps))
mean_steps = mean(dta$steps, na.rm = TRUE)
median_steps = median(dta$steps,na.rm = TRUE)
```
The number of Steps is highly skewed towards 0, 
with a **median of `r median_steps`** and **mean of `r mean_steps`**

## What is the average daily activity pattern?
```{r daily activity pattern}
means_per_interval = aggregate(steps ~ time, dta, mean)
with(means_per_interval, 
     plot(x = time,
          y = steps,
          type = "l",
          ylab = "Mean steps",
          main = "Daily patterns of activity"
          )
     )

max_mean = max(means_per_interval$steps) # Variable reported below
max_ind = which(means_per_interval$steps == max_mean)
max_interval = means_per_interval$time[max_ind]
max_interval = format(max_interval,"%H:%M") # Variable reported below
```

The maximum mean number of steps found in any interval (`r max_mean`)
is found in the interval starting at **`r max_interval`**.

## Imputing missing values

#### Report number of missing values
```{r Missing values}
count_of_missing_values = sum(is.na(dta$steps)) # Variable reported below

```

The total number of missing values is **`r count_of_missing_values`**.

#### Strategy for filling in missing values
The goal for filling missing values is to minimize the effect of new values on the overall structure.  The simplest method that has some chance of doing this is one that uses the mean activity at a particular time of day to replace values that are missing at that time of day (i.e. we assume that variation caused by time of day is greater than variation driven by the date).

#### Use strategy and create new dataset
```{r Perform imputation}

na_idx = is.na(dta$steps)
na_time_list = dta$time[na_idx]
matches = match(na_time_list,means_per_interval$time)
dta_imputed = dta
dta_imputed$steps[na_idx] = means_per_interval$steps[matches]
```

The new dataset, *dta_imputed*, has been created.

#### How does imputation affect the results?
```{r Imputation results}
# Count steps per day
steps_per_day     = aggregate(steps ~ date, dta,         sum)
steps_per_day_imp = aggregate(steps ~ date, dta_imputed, sum)

# Histograms
with(steps_per_day,hist(steps,breaks=20,main="Histogram of steps (original)"))
with(steps_per_day_imp,hist(steps,breaks=20,main="Histogram of steps (imputed)"))

#Table comparing results
means   = c(Original = mean(      steps_per_day$steps), 
            Imputed  = mean(  steps_per_day_imp$steps))
medians = c(Original = median(    steps_per_day$steps), 
            Imputed  = median(steps_per_day_imp$steps))
means = as.data.frame(means)
medians = as.data.frame(medians)
tbl = cbind(means,medians)
tbl
```

This form of imputation has minimal effect on the "center" because missing values are in entire days -- effectively in imputation increases the number of "average" days. Side effects for the choice of imputation method on the total number of steps in a day include:

- The mean remains unchanged.
- Days filled with NA are replaced with "average" days (increasing the number of such days)
- The median is driven closer to the mean.
- The variance (not displayed) is reduced.
- The shape of the histogram is changed except for the rectangle containing the mean, which increases in size.

The size of these effects are dependent on the number of missing values. This analysis also assumes that entire days are missing, not partial days.

## Are there differences in activity patterns between weekdays and weekends?
